---
title: "Promise Early Education"
author: "Fred, Soren, Katie K"
output: html_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(dplyr)
```

## 1. Introduction

Our data comes from Promise Early Education, a Head Start and Early Head Start program that works with vulnerable children to encourage educational growth and measures their development in six key areas: social/emotional, physical, and language development, as well literacy, cognitive, and mathematics knowledge and skills. These datasets measure scores in these various domains for children aged 0-5 and determines whether they are meeting expectations based on the expected scores for each age range. The variables include bottom expected scores, top expected scores, # of children, average score, # of children below score range, % of children below score range, # of children in score range, % of children in score range, the domain tested, age of children, the season, and the year.

We are hoping to explore the various trends seen within the promise early education, particularly the effect that the COVID-19 pandemic had on the several categories of child development in the years that followed it.


## 2. Data


```{r data-cleanup}
library(readxl)


pre_covid_fall_2018_19 <- read_excel("../data/pre_covid.xlsx", 
    sheet = "3-4", range = "A3:I81") |>
  janitor::clean_names()
  

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  rename(`category` = `x1`)

pre_covid_fall_2018_19[1, 1] <- "Social-Emotional"


pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  filter(!category %in% c("Threes SE", "Threes")) 

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

pre_covid_fall_2018_19$percent_meeting_exceeding <- as.numeric(pre_covid_fall_2018_19$percent_meeting_exceeding)

pre_covid_fall_2018_19_mean <- pre_covid_fall_2018_19 |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)

pre_covid_fall_2018_19_mean

#attempt to make a bar plot for %meeting/exceeding averages between each category
#ggplot(pre_covid_fall_2018_19_mean, aes(x=percent_meeting_exceeding)) +
 # geom_bar()+
  #facet_wrap(~category)


glimpse(pre_covid_fall_2018_19)

```


## 3. Data analysis plan
We are going to be exploring patterns before and after COVID-19 to visualize the effect that the COVID-19 pandemic has had on the several categories of child development. One variable that is going to be useful in this is,% Meeting/Exceeding, to see if these numbers differed before and after COVID-19. We intend to create a scatter plot, and use faceting since they are many different times we are comparing.  We have so much data, so we don't think we need additional data, but we are going to need to tidy up the data to make it easier to analyze.



## 4. Data Ethics Review

The data ethics review section will be introduced in a separate class and is not part of the original proposal deadline.

