---
title: "Project memo"
author: "Soren, Katie, & Fred"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

## Data Clean Up Steps for Overall Data

### Step 1: Clean 2018-19 Data 

```{r cleanup data-18-19}
library(readxl)

library(janitor)
######FALL DATA###########

pre_covid_fall_2018_19 <- read_excel("../data/pre_covid.xlsx", 
    sheet = "3-4", range = "A3:I81") |>
  janitor::clean_names()
  

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  rename(`category` = `x1`)

pre_covid_fall_2018_19[1, 1] <- "Social-Emotional"


pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  filter(!category %in% c("Threes SE", "Threes")) 

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

pre_covid_fall_2018_19 <- pre_covid_fall_2018_19 |>
  mutate(season = "Fall") |>
  mutate(year = "18-19")

pre_covid_fall_2018_19$percent_meeting_exceeding <- as.numeric(pre_covid_fall_2018_19$percent_meeting_exceeding)

pre_covid_fall_2018_19_mean <- pre_covid_fall_2018_19 |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)

pre_covid_fall_2018_19_mean


######WINTER DATA###########

pre_covid_winter_2018_19 <- read_excel("../data/pre_covid.xlsx", 
    sheet = "3-4", range = "J3:O80")

expectations_18_19 <- read_excel("../data/pre_covid.xlsx", 
    sheet = "3-4", range = "A3:C80")


pre_covid_winter_2018_19 <- cbind(expectations_18_19, pre_covid_winter_2018_19) |>
    janitor::clean_names()

pre_covid_winter_2018_19 <- pre_covid_winter_2018_19 |>
  rename(`category` = `x1`)

pre_covid_winter_2018_19[1, 1] <- "Social-Emotional"


pre_covid_winter_2018_19 <- pre_covid_winter_2018_19 |>
  filter(!category %in% c("Threes SE", "Threes")) 

pre_covid_winter_2018_19 <- pre_covid_winter_2018_19 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

pre_covid_winter_2018_19 <- pre_covid_winter_2018_19 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

pre_covid_winter_2018_19 <- pre_covid_winter_2018_19 |>
  mutate(season = "Winter") |>
   mutate(year = "18-19")

pre_covid_winter_2018_19$percent_meeting_exceeding <- as.numeric(pre_covid_winter_2018_19$percent_meeting_exceeding)

pre_covid_winter_2018_19_mean <- pre_covid_winter_2018_19 |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)


######SPRING DATA###########

pre_covid_spring_2018_19 <- read_excel("../data/pre_covid.xlsx", 
    sheet = "3-4", range = "P3:U80")

pre_covid_spring_2018_19 <- cbind(expectations_18_19, pre_covid_spring_2018_19) |>
    janitor::clean_names()


pre_covid_spring_2018_19 <- pre_covid_spring_2018_19 |>
  rename(`category` = `x1`)

pre_covid_spring_2018_19[1, 1] <- "Social-Emotional"


pre_covid_spring_2018_19 <- pre_covid_spring_2018_19 |>
  filter(!category %in% c("Threes SE", "Threes")) 

pre_covid_spring_2018_19 <- pre_covid_spring_2018_19 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

pre_covid_spring_2018_19 <- pre_covid_spring_2018_19 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

pre_covid_spring_2018_19 <- pre_covid_spring_2018_19 |>
  mutate(season = "Spring") |>
   mutate(year = "18-19")

pre_covid_spring_2018_19$percent_meeting_exceeding <- as.numeric(pre_covid_spring_2018_19$percent_meeting_exceeding)

####MERGE PRE-COVID DATA######

pre_covid <- rbind(pre_covid_fall_2018_19, pre_covid_winter_2018_19, pre_covid_spring_2018_19) |>
  mutate(season = fct_relevel(season, "Fall", "Winter", "Spring"))

#should we make a variable for age, since the data is split between age groups? or just choose 3-4 year olds as our #data set?
#pre_covid <- pre_covid |>
 # mutate(children_age = "3-4")

pre_covid_spring_2018_19_mean <- pre_covid_spring_2018_19 |>
  group_by(category) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)
```

### Step 2: Clean 20-21 Data

```{r cleanup-data-20-21}

#####FALL DATA########

fall_2020_21 <- read_excel("../data/2020_21_HS.xlsx", 
    sheet = "HS RAW ALL", range = "A4:K50") |>
    janitor::clean_names()

fall_2020_21 <- fall_2020_21 |>
  rename(`category` = `x1`)

fall_2020_21[1, 1] <- "Social-Emotional"


fall_2020_21 <- fall_2020_21 |>
  filter(!str_detect(category, "TOTAL")) 

fall_2020_21 <- fall_2020_21 |>
 mutate(category = case_when(str_detect(category, pattern = "Language|Mathematics|Social-Emotional|Cognitive|Physical|Literacy") ~ category, TRUE ~ NA))
 
 
 # mutate(category = na_if(category, category == str_detect("Yrs")))
  # mutate(category = case_when(category == str_detect("Lang") ~ "Language",
#category == str_detect(...),
#TRUE ~ "No category"))

fall_2020_21 <- fall_2020_21 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

fall_2020_21 <- fall_2020_21 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

fall_2020_21 <- fall_2020_21 |>
mutate(across(2:11, as.numeric))

fall_2020_21[is.na(fall_2020_21)] <- 0

#fall_2020_21$percent_meeting <- as.numeric(fall_2020_21$percent_meeting) 
 #fall_2020_21$percent_exceeding <- as.numeric(fall_2020_21$percent_exceeding)
 #fall_2020_21$number_meeting <- as.numeric(fall_2020_21$number_meeting)
 #fall_2020_21$number_exceeding <- as.numeric(fall_2020_21$number_exceeding)

fall_2020_21 <- fall_2020_21 |>
  mutate(season = "Fall") |>
   mutate(year = "20-21") 

######WINTER DATA##########

winter_2020_21 <- read_excel("../data/2020_21_HS.xlsx", 
    sheet = "HS RAW ALL", range = "N4:U50") |>
    janitor::clean_names()

expectations_20_21 <- read_excel("../data/2020_21_HS.xlsx", 
    sheet = "HS RAW ALL", range = "A4:C50")

winter_2020_21 <- cbind(expectations_20_21, winter_2020_21) |>
      janitor::clean_names()
  

winter_2020_21 <- winter_2020_21 |>
  rename(`category` = `x1`)

winter_2020_21[1, 1] <- "Social-Emotional"

winter_2020_21 <- winter_2020_21 |>
  filter(!str_detect(category, "TOTAL")) 

winter_2020_21 <- winter_2020_21 |>
 mutate(category = case_when(str_detect(category, pattern = "Language|Mathematics|Social-Emotional|Cognitive|Physical|Literacy") ~ category, TRUE ~ NA))
 

winter_2020_21 <- winter_2020_21 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

winter_2020_21 <- winter_2020_21 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))


winter_2020_21$percent_meeting <- as.numeric(winter_2020_21$percent_meeting) 
winter_2020_21$percent_exceeding <- as.numeric(winter_2020_21$percent_exceeding)
winter_2020_21$number_meeting <- as.numeric(winter_2020_21$number_meeting)
winter_2020_21$number_exceeding <- as.numeric(winter_2020_21$number_exceeding)

winter_2020_21 <- winter_2020_21 |>
  mutate(season = "Winter") |>
   mutate(year = "20-21")

#####SPRING DATA##########

spring_2020_21 <- read_excel("../data/2020_21_HS.xlsx", 
    sheet = "HS RAW ALL", range = "X4:AE50") |>
    janitor::clean_names()

spring_2020_21 <- cbind(expectations_20_21, spring_2020_21) |>
      janitor::clean_names()
  

spring_2020_21 <- spring_2020_21 |>
  rename(`category` = `x1`)

spring_2020_21[1, 1] <- "Social-Emotional"

spring_2020_21 <- spring_2020_21 |>
  filter(!str_detect(category, "TOTAL")) 

spring_2020_21 <- spring_2020_21 |>
 mutate(category = case_when(str_detect(category, pattern = "Language|Mathematics|Social-Emotional|Cognitive|Physical|Literacy") ~ category, TRUE ~ NA))
 
 
 # mutate(category = na_if(category, category == str_detect("Yrs")))
  # mutate(category = case_when(category == str_detect("Lang") ~ "Language",
#category == str_detect(...),
#TRUE ~ "No category"))

spring_2020_21 <- spring_2020_21 |>
  fill(category, .direction = "down") |>
  drop_na(bottom)

spring_2020_21 <- spring_2020_21 |>
  filter(!bottom %in% c("Widely Held Expectations", "Bottom"))

spring_2020_21 <- spring_2020_21 |>
mutate(across(2:11, as.numeric))

spring_2020_21$percent_meeting <- as.numeric(spring_2020_21$percent_meeting) 
 spring_2020_21$percent_exceeding <- as.numeric(spring_2020_21$percent_exceeding)
 spring_2020_21$number_meeting <- as.numeric(spring_2020_21$number_meeting)
 spring_2020_21$number_exceeding <- as.numeric(spring_2020_21$number_exceeding)

spring_2020_21 <- spring_2020_21 |>
  mutate(season = "Spring") |>
   mutate(year = "20-21")


```
### Step 3: Merging 18-19 & 20-21 data

```{r merge-pre-during-data}

covid_2020_21 <- rbind(fall_2020_21, winter_2020_21, spring_2020_21) |>
  mutate(season = fct_relevel(season, "Fall", "Winter", "Spring"))

covid_2020_21 <- covid_2020_21 |>
mutate(across(2:11, as.numeric))

covid_2020_21[is.na(covid_2020_21)] <- 0

covid_2020_21 <- covid_2020_21 |>
  mutate(number_meeting_exceeding = number_meeting + number_exceeding) |>
mutate(percent_meeting_exceeding = percent_meeting + percent_exceeding) |>
  select(-season, season) |>
  select(-year, year)

covid_2020_21 <- covid_2020_21 |>
  select(-(number_meeting:percent_exceeding))

scores <- rbind(pre_covid, covid_2020_21)
```

### Step 4: Clean 22-23 Data


```{r post-covid-cleanup}
library(readxl)

library(janitor)

####### FALL DATA ########

post_covid <- read_excel("../data/post_covid.xlsx",
  range = "A3:K65") |>
  janitor::clean_names()

post_covid <- post_covid |>
  rename(`category` = `x1`)


post_covid <- post_covid %>% filter(!if_all(-1, is.na)) #removes rows of all NA
post_covid <- post_covid[, -c(2,3)]
post_covid <- post_covid %>% filter(!is.na(average))


post_covid[1:5, 1] <- "Social-Emotional"
post_covid[6:10, 1] <- "Physical"
post_covid[11:15, 1] <- "Language"
post_covid[16:20, 1] <- "Cognitive"
post_covid[21:25, 1] <- "Literacy"
post_covid[26:30, 1] <- "Mathematics"

#view(post_covid)
fall <- post_covid[-c(5, 10, 15, 20, 25, 30), ]
#view(fall)

#### WINTER AND SPRING DATA #########
post_covid_win_spr <- read_excel("../data/post_covid.xlsx",
  range = "M3:AE65") |>
  janitor::clean_names()

post_covid_win_spr <- post_covid_win_spr |>
  rename(`category` = `x1`) |>
  rename(`category_1` = `x11`)


post_covid_win_spr <- post_covid_win_spr %>% filter(!if_all(-1, is.na)) #removes rows of all NA
post_covid_win_spr <- post_covid_win_spr %>% filter(!is.na(average_3))


post_covid_win_spr[1:5, 1] <- "Social-Emotional"
post_covid_win_spr[6:10, 1] <- "Physical"
post_covid_win_spr[11:15, 1] <- "Language"
post_covid_win_spr[16:20, 1] <- "Cognitive"
post_covid_win_spr[21:25, 1] <- "Literacy"
post_covid_win_spr[26:30, 1] <- "Mathematics"

post_covid_win_spr[1:5, 11][is.na(post_covid_win_spr[1:5, 10])] <- "Social-Emotional"
post_covid_win_spr[6:10, 11][is.na(post_covid_win_spr[6:10, 10])] <- "Physical"
post_covid_win_spr[11:15, 11][is.na(post_covid_win_spr[11:15, 10])] <- "Language"
post_covid_win_spr[16:20, 11][is.na(post_covid_win_spr[16:20, 10])] <- "Cognitive"
post_covid_win_spr[21:25, 11][is.na(post_covid_win_spr[21:25, 10])] <- "Literacy"
post_covid_win_spr[26:30, 11][is.na(post_covid_win_spr[26:30, 10])] <- "Mathematics"



post_covid_win_spr <- post_covid_win_spr[-c(5, 10, 15, 20, 25, 30),]
post_covid_win_spr <- post_covid_win_spr[, -c(10)]


```



### Merge 22-23 Data

```{r merging-final-data}
#create season column
fall <- fall %>%
  mutate(season = "Fall")
#view(fall)

winter <- post_covid_win_spr[, 1:9]  # Columns 1 to 10
winter <- winter %>%
  mutate(season = "Winter")

spring <- post_covid_win_spr[, 10:ncol(post_covid_win_spr)]
spring <- spring %>%
  mutate(season = "Spring")

#rename columns
colnames(winter) <- c("category", "number_children", "average", "number_below", "percent_below", 
                                  "number_meeting", "percent_meeting", "number_exceeding", "percent_exceeding", "season")
#view(winter)

colnames(spring) <- c("category", "number_children", "average", "number_below", "percent_below", 
                                  "number_meeting", "percent_meeting", "number_exceeding", "percent_exceeding", "season")
#view(spring)

#merge dfs

post_covid_final  <- rbind(fall, winter, spring)

#view(post_covid_final)
              
post_covid_final <- post_covid_final |>
mutate(across(2:9, as.numeric))          

#post_covid_final[is.na(post_covid_final)] <- 0

post_covid_final <- post_covid_final |>
  mutate(number_meeting_exceeding = number_meeting + number_exceeding) |>
mutate(percent_meeting_exceeding = percent_meeting + percent_exceeding) |>
 mutate(year = "22-23") |>
  select(-season, season) |>
  select(-year, year)

post_covid_final <- post_covid_final |>
  select(-(number_meeting:percent_exceeding))

##### FINAL MERGE #######

#scores<-scores |>
  #select(-(bottom:top))


#scores <- rbind(scores, post_covid_final)

#scores[is.na(scores)] <- 0
scores<-scores |>
  select(-(bottom:top))


scores <- rbind(scores, post_covid_final)

#scores[is.na(scores)] <- 0

scores_mean <- scores |>
  group_by(category, season,year) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)
```


## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```



### Plot 1: Mean Scores Plot

#### Data cleanup steps specific to plot 1

These data cleaning sections are optional and depend on if you have some data cleaning steps specific to a particular plot

```{r data-cleaning-plot1}

#pre_covid[is.na(pre_covid)] <- 0


pre_covid_mean <- pre_covid |>
  group_by(category, season) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)



```


#### Final Plot 1

```{r mean-plot-18-19,  fig.alt = c("Bar graphs faceted by test category that show percentage of children at or above expectations over Fall, Winter, and Spring for the year '18-'19. For all categories, Cognitive, Language, Literacy, Mathematics, Physical, and Social-Emotional, there is an upward trend as the year goes on.")}


ggplot(pre_covid_mean, aes(x = season, y = percent_meeting_exceeding, fill = category)) +
  geom_col() +
  labs(
    y = "Percent Meeting/Exceeding",
    x = "Season",
    title = "Percent of Children Meeting/Exceeding Test Score Expectations in Each Category",
    subtitle = "For Year 2018-19",
    fill = "Category")  +
  ylim(0, 100) +
  facet_wrap(~category) +
  guides(fill = "none")+
  theme(
  plot.background = element_rect(fill="white")
  ) +
  scale_fill_viridis_d() 

ggsave("18_19_graphs.png", width = 6, height = 4, dpi = 300)


```


### Plot 2: Academic Performance by Category Across Seasons Line Graph: Pre, During, Post COVID trends

#### Data cleanup steps specific to plot 2

These data cleaning sections are optional and depend on if you have some data cleaning steps specific to a particular plot
```{r cleaning-plot2}
score_fallspring <- scores |>
  filter(season != "Winter") |>
  group_by(category, season, year) |>
  summarize(score_mean = mean(percent_meeting_exceeding))


score_fallspring_wide <- score_fallspring |>
  pivot_wider(names_from = season,
              values_from = score_mean) |>
  mutate(percent_increase = Spring - Fall) #|>
 # pivot_wider(names_from = year,
  #            values_from = percent_increase)


##score_fallspring_wide |>
 ## summarize(percent_increase,-category)

#result <- score_fallspring_wide[seq(1, nrow(score_fallspring_wide), 2),] - score_fallspring_wide[seq(2, nrow(score_fallspring_wide), 2),] 

#scores<-scores |>
#select(-(bottom:top))
scores <- scores |> select(-any_of(c("bottom", "top")))




scores <- rbind(scores, post_covid_final)

#scores[is.na(scores)] <- 0
scores <- scores |> mutate(across(where(is.numeric), ~replace_na(., 0)))



scores_mean <- scores |>
  group_by(category, season,year) |>
  summarize_at(vars("percent_meeting_exceeding"), mean, na.rm = TRUE)

```

#### Final Plot 2
```{r covid-line-plot, fig.alt = c("Three line plots faceted by year that show the percentage of children that scored at or above expectations, with one line for each category. There is an upward trend for all categories in the years `18-`19 and `20-`21")}



ggplot(scores_mean, aes(x = season, y = percent_meeting_exceeding, group = category, color = category)) +
  geom_line() + 
  geom_point() +
  labs(
    title = "Academic Performance by Category Across Seasons",
    subtitle = "Grouped by COVID-19 Era: Pre, During, and Post",
    x = "Season",
    y = "Percent Meeting/Exceeding",
    color = "Category",
  ) +
  facet_wrap(~ year, ncol = 3) +
  scale_color_brewer(palette = "RdBu")+ 
  theme_minimal() +
  theme(
  legend.position = "bottom",
  legend.direction = "horizontal",
plot.background = element_rect(fill="white")  
)

ggsave("covid_line_plot.png", width = 8.5, height = 5, dpi = 300)
```


### Plot 3: Heatmap of Student Performance Across COVID Periods and Seasons

```{r category-heatmap, fig.alt = c("Heat map that shows the percentage of children that scored below expectation or meeting/exceeding expectations in test scores for different categories across seasons and years. All seasons across the year '22-'23 generally have lower percentage of children at or above the threshold")}

scores_mean <- scores_mean %>%
  mutate(period_season = paste(year, season, sep = "-"))

scores_mean <- scores_mean %>%
  mutate(
    percent_below = 100 - percent_meeting_exceeding
  )

scores_long <- scores_mean %>%
  pivot_longer(
    cols = c(percent_meeting_exceeding, percent_below),
    names_to = "type",
    values_to = "percent"
  )


scores_long_combined <- scores_long %>%
  mutate(status = ifelse(type == "percent_meeting_exceeding", "Meeting/Exceeding", "Below Expectations"))

scores_meeting <- scores_long_combined %>% 
  filter(status == "Meeting/Exceeding")

ggplot(scores_meeting, aes(x = period_season, y = category, fill = percent)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightyellow", high = "darkred") +
  labs(
    title = "Student Performance: Percent Meeting/Exceeding Expectations",
    x = "Year - Season",
    y = "Category",
    fill = "% Meeting/Exceeding"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.background = element_rect(fill="white")
  )


ggsave("performance_heatmap.png", width = 7.5, height = 4.5, dpi = 300)






```
